---
title: "Description of the scBio deconvolution method"
output:
  html_document:
    df_print: paged
---

This is a personal document to demonstrate the key functions used in the paper with corresponding code snippets.    
<https://www.nature.com/articles/s41592-019-0355-5>

There are four main inputs that are required: 
1) scRNA count matrix (reference)  
2) a cell-state map derived from the scRNA data (e.g.tSNE/UMAP projection that has been pre-defined)  
3) cell labels (reference cell labels. e.g. cell clusters (B cells, T cells, etc))  
4) a bulk RNA-seq data it wants to deconvolve  

This program runs multiple iterations and average the resulting values for a final decovolution result.  
So the result is a map of relative abundance of cell states annd types imputed from the input bulk RNA-seq data.  
Per iteration, this program samples random cells representing every cell clusters and states.
The number of iterations and randomly selected cells per iteration are pre-computed before it runs support vector regression for actual deconvolution. 

So the first step is to define the number of iterations and assign randomly selected cells per iteration.
A main function that does this is **choseCellsForRuns()**

It starts by firstly defining a grid.   
An 'initialGrids' object is a 'list' object containing cells (names) sorted into 'k' number of bins per cell cluster

```{r}
#1. This is the first part of the choseCellsForRuns()

initialGrids = lapply(unique(SCLabels), function(currCluster){
  clusterIndexes = which(SCLabels==currCluster)
  nbins = max(k,length(clusterIndexes)/neighborhoodSize)
  
  currXY = SCCellSpace[clusterIndexes,] # Dim1 and Dim2 values of a cell cluster
  ch <- grDevices::chull(currXY) # Stores indices of a convex hull (x and y coordinates that enclose the geometric shape)
  coords <- currXY[c(ch, ch[1]), ] # Coordinates of the indices
   
  poly = sp::SpatialPolygons(list(sp::Polygons(list(sp::Polygon(coords)), "x"))) # A polygon object
  grid <- raster::raster(raster::extent(poly), nrows = ceiling(sqrt(nbins)), ncols= ceiling(sqrt(nbins))) # Extent function returns xmin, xmax, ymin, ymax (coordinates)
  sp::proj4string(grid)<-sp::proj4string(poly) # Grid object has a grid with 3 X 3 dimension enclosed by the coordinates defined above 
  cellLocationOnGrid = raster::cellFromXY(grid, currXY) 
  # Assign cells within the cluster to this grid (Since it is a 3 X 3 grid, we are trying to sort the cells into 9 bins, but not necessarily filling all the 9 bins)
  tab <- table(cellLocationOnGrid)
  grid[as.numeric(names(tab))] <- tab
  
  list(grid = grid, clusterIndexes = clusterIndexes, cellLocationOnGrid = cellLocationOnGrid, nFullbins = length(tab), maxBinSize = max(tab))
}) 
# Sorting the cells into 'k' number of bins is important for the downstream process where we sample 'k' cells from these bins 
# These ensures the program samples cells equally across the cell-state space

numOfRuns = ceiling(minSelection*max(unlist(lapply(initialGrids,function(clusterData){ clusterData$nFullbins * clusterData$maxBinSize  }))) / k) 
# Number of iterations depend on the a) the number of bins, b) the number of cells per bin, and c) the number of unique cell labels

```

The second step is to ramdoly sample from these grids per iteration  
The goal here is to sample only one cell from each bin per cell label

``` {r}
#2. This is the second part of the choseCellsForRuns()

meanDistMatrix = rep(1,length(refNames))
chosenCellList = lapply(1:numOfRuns, function(runNum){
  chosenCells = as.numeric(unlist(lapply(unique(refNames),function(currCluster){
    initialGrid = initialGrids[[which(unique(refNames)==currCluster)]]
    clusterIndexes = initialGrid$clusterIndexes
    grid = initialGrid$grid
    cellLocationOnGrid = initialGrid$cellLocationOnGrid
    kToUse = k
    if(k>length(which(!is.na(grid[])))){
      kToUse = length(which(!is.na(grid[])))
      }
    gridCellsToUse = sample(which(!is.na(grid[])),kToUse,replace = F)
    chosenCellsForCluster = clusterIndexes[unlist(lapply(gridCellsToUse, function(currCell){
      chosenCell = which(cellLocationOnGrid==currCell)
      if(length(chosenCell)>1){
        chosenCell = sample(chosenCell,1,prob = meanDistMatrix[clusterIndexes[chosenCell]])
        }
      chosenCell
      }))]
    chosenCellsForCluster
    })))
  cellsToReduce = chosenCells[which(meanDistMatrix[chosenCells]>minValueToReduceTo)]
  meanDistMatrix[cellsToReduce] <<- meanDistMatrix[cellsToReduce]/10
  chosenCells
  })

```

The last step of this function is to derive 10 most closely related cells for each cell.  
For example, if there are 10 cell types and 10 bins representing the cell states within a cell type, then one iteration is comprised of 1000 cells.  

```{r}
#3. This is the last part of the choseCellsForRuns()

chosenNeigList = lapply(1:length(SCLabels),function(cellIndex){ # Select one cell
  selectedCellType = SCLabels[cellIndex]  
  selectedCellIndexes = which(SCLabels == selectedCellType)
  cellXY = SCCellSpace[cellIndex,] # Selected cell's Dim1 and Dim2 information
  cellDist = fields::rdist(t(as.matrix(cellXY)),SCCellSpace[selectedCellIndexes,]) # Distance measure between the selected cell and other cells in the same cluster
  chosenRepeats = order(as.numeric(cellDist),decreasing = F)[1:neighborhoodSize] # Top 10 similar cells
  chosenRepeats = chosenRepeats[!is.na(chosenRepeats)]
  selectedCellIndexes[chosenRepeats]
})

cellselection = list(chosenCellList = chosenCellList, chosenNeigList = chosenNeigList, numOfRuns = numOfRuns)

# In the main function, it will use the sampled cells per cluster ( k * # of clusters / run) defined above and and 10 other similar cells (neighbor cells) per sampled cells
# Also, sample reduced number of genes to use per iteration 

```

These randomly chosen cells per iteration are then used as inputs for support vector regression.  

Selecting genes to use per iteration:  

For each iteration, the program randomly select genes to use (using 'sample' function).  
Then, these gene are further filtered by using 1) ANOVA to filter out genes that show high intra-group variation ('group' here is the cell type) and 2) custom kappa function.    
Finally, a support vector regression is used to impute the cell states and types per iteration using these randomly selected genes.
